TAPIR: Threaded Application Programming on Intermittent Resources

### 2017-05-12 paper.tex
Fixed code listings
Added figures for interrupt control flow and the threads array
Spelling fixes

In this paper we introduce the \textit{Threaded Application Programming on Intermittent Resources} 
(TAPIR) model to expand the current intermittent programming model \textit{Chain} to 
include general multiprogramming. TAPIR allows the range of possible programs for 
intermittent devices to include more general computation functionality while leaving the
memory and energy overhead low enough that intermittent computation on those workloads 
still guarantees forward progress. While nonvolatile memory usage will increase per thread
scheduled, certain optimizations are possible to reduce memory overhead overall. Additionally,
context switching through our scheduler task is a low-instruction count static overhead that is
easily accounted for in energy calculations. Through the design of TAPIR we show that general
purpose threaded applications are feasible on an intermittently powered device.

\end{abstract}

@@

Our multithreaded model is build on two existing programming models, DINO and
Chain.  An initial intermittent programming model DINO breaks intermittent
programs into tasks, which creates automatic checkpointing locations during
exeuction \cite{Dino}. DINO uses a checkpointing and recovery model that tracks
both volatile and nonvolatile memory.  The DINO execution model executes all
tasks transactionally, leading to atomic task semantics and does not permit
consistency violations in nonvolatile state \cite{Dino}. Chain, a more recent
iteration of the concepts in DINO, uses volatile state rarely, if ever, and uses
task boundaries called channels to store most used data in nonvolatile memory
\cite{Chain}. At compile time, Chain allocates the necessary channels between
tasks by parsing channel read and write statements in order to match up aligning
tasks. All inputs and outputs to each task, as well as necessary data
structures, are stored in a channel for a respective task. A channel may only be
modified by one task, and only one task may run at a time, which mitigates time
of check, time of use bugs (TOCTOU) in intermittence.  The one exception is a
self-channel for recursive code, which is double buffered.  Keeping all
modifiable data in nonvolatile memory that is not open to easy rewrite makes
these tasks idempotent across reboots.  Because of continuous reboots, many
operations necessarily will be executed more than once, and correctness is
compromised if those operations are not guaranteed to have the same outcome by
the programming model after each reboot.

@@

\section{Related Work}
In addition to the division of a program into tasks, as described above, prior work
explores several different strategies for preserving state across reboots. The approaches
range from fully independent of the programmer~\cite{ratchet, dewdrop} at one extreme, to programmer
defined task boundaries~\cite{Dino} at the other.  Compiler driven approaches such as the
method used in ~\cite{Mementos} allow the programmer to specify high level behavior of the
checkpointing system. A key similarity among the prior work is it
does not lend itself well to multi-programming let
alone multi-threading. In fact, many of these works state the assumption that the device
hardware cannot support a conventional OS, and thus focus on single threaded programs.
Despite the availability of (relatively) inexpensive nonvolatile memory technology such as
FRAM~\cite{quickrecall}, the the energy and storage overhead of checkpointing multiple
thread states would limit the maximum per-thread checkpoint size.

@@

While prior work has stopped short of suggesting an operating system for intermittently
powered devices, a number of different works touch on the need for these devices to
perform multiple tasks in the same workload. Buettner, Greenstein and Wetherall motivate
their runtime, Dewdrop, based on difference in energy requirements for the tasks that a
CRFID needs to perform~\cite{dewdrop}. The energy aware memory mapping in ~\cite{Aware}
discusses the challenge of handling nondeterministic interrupts received by intermittent
sensing platforms when a peripheral must be serviced immediately and control flow deviates
from single threaded execution. From a hardware standpoint, the federated energy scheme in
~\cite{ufop} provides a platform for removing the energy dependence between an MCU and its
peripherals. Federated energy also simplifies a programmer's reasoning about reading from
a single peripheral. However, federating energy does not help a programmer describe how to
handle {\em multiple} peripherals. Consider a listing in
~\cite{ufop} reporoduced in Figure ~\ref{fig:copy}, and imagine the level of understanding
of the hardware behavior a programmer would have to employ to, say, read one sensor if it
is available but read from and proceed to process data from a different sensor if it is not.

@@

The wireless sensor networks community has explored multiple approaches for handling
multitasking on resource limited devices. The designs tend to provide a combination of
timer and event driven context switching as well as a simple programming interface.
TinyOS uses a simple FIFO scheduler to handle event based context switches~\cite{tiny}.
Nano-RK implements priority based preemptive multitasking where task priorities are
statically defined~\cite{nano}. Contiki uses proto-theads, stackless program threads, to
offer a model between nano-RK and tinyOS. The Contiki implements "cooperative
multithreading" such that events can trigger a context switch, but multithreading may also
be timer based~\cite{contiki}. The goal of multitasking interfaces in wireless sensor
networks is to provide a tradeoff between network lifetime and quality of service. This
differs from the intermittent programming model where the challenge is not system
lifetime, but system functionality based on energy availability.

@@

\section{System Implementation}
TAPIR has a round robin scheduling algorithm. At the end of a task, the compiler
inserts a jump back to the scheduler task, where the next task in the list is
the one to be executed. An alternate scheduling model that was considered was
a context switch on reboot instead of at the end of each task. The main drawback
of using a reboot-based scheduler is that, because power failure is the common
case, there is a high chance of reducing or halting forward progress altogether
because one thread holds a lock that all of the other threads need but cannot
complete. For that reason the design specifies end of task, so each task has
an opportunity to use a shared resource and release any locks before a power-
intensive context switch occurs. Additionally, a round robin scheduler was
chosen for this design because it is simple to implement and uses only one index counter,
very little power or memory overhead as compared to other scheduling algorithms. It is
critical for multithreading in Chain to not add significant power and memory
overhead in scheduling over conventional Chain so that the workload can
make as much forward progress as possible on one charge cycle. While out of
scope for this project, future work may include a priority scheduler to decrease
latency for certain threads due to context switching for I/O or other input latency
sensive workloads.

When designing threading in Chain, the shared memory model is a main indicator
of the threaded execution model. The two most probable options when designing
TAPIR were a SIMD execution model of \textit{Thread Pools}, or a MIMD
traditional multiprogramming model. The Thread Pool design makes use of task
code duplication by explicitly forking a number of chain programs, all of which
run simultaneously on the scheduler on arrayed data. This option is useful
multiple data streams from I/O in embedded applications that need to be
processed before new data overrides the I/O device. Additionally, the Thread
Pool model makes sharing memory more regular by using multiple copies of the
same thread. However, this option is not as general purpose. The TAPIR design
ultimately settled on a multiprogramming model, because this allows for the
programmer to explictly program unique chain programs with shared or discrete
data structures on the same system. However, this execution model requires a
more structured locking mechanism to avoid RMW during execution.

SIMD execution on an intermittent system relies on nonvolatile memory instead
of independent registers for each thread. In Chain's case the implementation
would rely on channels for each thread running in the system. In the MIMD
execution model channels could potentially be re-assigned to already allocated
channels because the threads may end at interleaving times. However, in the
SIMD model it is expected that each identical thread is at a comperable point
in execution to the other running threads, and so each thread that may be
run necessitates a new channel allocation. This could lead to an unavoidable
hard limit on the number of threads that can run in an applicatioin which in
turn limits potential applications that are useful on this intermittent device.

@@

\subsection{Locking Primitive}
To improve a programmer's control of shared data between threads, and to reduce
the complexity of accessing data that may be touched by many different tasks we
implement a simple mutex object that can be added to a channel. The mutex is
designed to be used in a standard \texttt{channel}, not a double buffered
\texttt{self\_channel}, so it must be able to tolerate a failure after a
read/write operation (i.e. test and set). To achieve idempotence, we add a
\texttt{thread\_id} to the \texttt{free} bit inside of each mutex. When
attempting to acquire a lock, a thread checks if its thread id is equal to the
lock holder, then it checks if the lock is free. If either condition is true,
the thread successfully acquires the lock, otherwise the \texttt{mutex\_lock}
function fails. Upon successfully acquiring the lock, a thread writes its id to
the mutex's holder id, and changes the \texttt{free} bit to \texttt{status =
held}. This scheme guarantees correctness because if a failure occurs after the
mutex is set but before the task commits, the task will read out the id of the
mutex holder before checking if it is free. Upon matchint its id with the holder
id, the task will proceed regardless of the fact that \texttt{status = held}.
%TODO Maybe add a figure here to make the sequence of read/write/failure clearer...
Further, a task must finsih executing before switching to another thread, so
there is no possiblity for an external thread to see a value that was not
committed by another task.

It is important to note that the mutex object can be modified idempotently, but
accesses to the data in the same channel as the mutex must adhere to channel
exclusion. In our example application, the mutex is contained within a user
defined \texttt{shared\_channel} structure that is analogous to a
\texttt{multicast\_channel} but abandons the naming convention that "restricts"
the channel to a static subset of tasks. Any task may read {\em or} or write to
the \texttt{shared\_channel}, but performing both would break correctness. In
the future, a compiler could be used to guarantee that a
\texttt{shared\_channel} is not read and written by the same task, but at
present we believe that the \texttt{mutex} and \texttt{shared\_channel}
interface is easier for programmers to use than \texttt{multicast\_channels}, or
a sepcialized double buffered channel when passing data between threads. It
maintains familiar semantics for programmers and does not incur the memory
overhead of double buffering.

@@

@@

\section{Implementation}
\subsection{Scheduler}
\begin{figure*}
\begin{minipage}[b]{.48\textwidth}
\begin{lstlisting}[language=C]
    typedef struct thread_t {
        unsigned thread_id;
        context_t context;
    } thread_t;

@@

    // List of free indicies in threads[]
    unsigned free_indicies[MAX_NUM_THREADS];
    unsigned free_indicies_size;

    // Index of current running thread
    // Stored in self channel
    unsigned current;
\end{lstlisting}
\caption{Scheduler Data Structures}\label{schDataStruct}
\end{minipage}\hfill
\end{figure*}

The data structures for our scheduler are listed in figure \ref{schDataStruct}.
Like the rest of chain, memory for data structures is allocated at compile
time. Thus, the limit on the maximum number of threads allowed in a system is
fixed arbitrarily - not ideal, but there isn't a better option while operating
inside chain. The \texttt{threads} array is stored in all-to-one channel, from
every task to the scheduler task. The \texttt{free\_indicies} array is stored
in a one-to-all channel, from the scheduler task to every task.

With these multicast channels, the scheduler task can act as a buffer between
multiple threads - arbitrary tasks write \texttt{threads[]} with calls to
\texttt{thread\_create} and \texttt{thread\_end} and tasks read from
\texttt{free\_indicies} through calls to thread create, maintaining
idempotency. Inside the scheduler task, we read and write \texttt{current},
so \texttt{current} is double buffered. The only issue is with
\texttt{thread\_init} since it is called from user-defined tasks and
initializes \texttt{current} - the same task that called \texttt{thread\_init}
could call \texttt{thread\_create}, and the \texttt{thread\_create} call
wouldn't see the update to the variable. Since \texttt{current} is always
initialized to the same value (0), we opt for manually swapping the buffer on
\texttt{current} so that other threads can see the update.

We keep an unsigned integer, \texttt{curr\_free\_index}, in volatile memory,
zeroed on every reboot, that acts as a current pointer in
\texttt{free\_indicies[]}. Every time the scheduler task runs,
\texttt{free\_indicies[]} is updated with the state of \texttt{threads[]},
based on whether or not the active bit is set for a given index. When
\texttt{thread\_create} is called, if \texttt{curr\_free\_index} is greater
than \texttt{free\_indicies\_size}, then \texttt{thread\_create} will fail in
POSIX fashion - the system does not have enough resources to support another
thread. Otherwise, \texttt{threads[free\_indicies[curr\_free\_index]]} is set
to the new thread. Upon power failure, \texttt{curr\_free\_index} is reset to
zero, so as long as the rest of the operations in the task are idempotent, so
will \texttt{thread\_create} calls - the same locations in \texttt{theads[]}
will be overwritten.

% TODO picture of array?

\subsection{Interrupts}
The primary issue with interrupts is the interrupt enable flags. The MSP430,
like many other embedded arhitetures has a master interrupt enable,
as well as specific, device-level interrupt enables. Both are the user's
responsibility, and need to be specified in the user-defined
\texttt{INIT\_FUNC}, as both will be cleared on reset. When an interrupt fires,
however, interrupts will be disabled by the CPU (again, on most architectures)
and if the device loses power immediately after an interrupt occurs, there will
be no record that the interrupt occured. This case is unaviodable on
intermittent power without explicit support from the power system - we reason
that if the user cared about every interrupt that fired and if missing one
interrupt would be catestrophic to the device's operation, then, arguably, that
device should not be intermittently powered.

Beyond this, there remains the issue of how to return to the interrupt task
with the interrupt enable set to the correct value if a power failure occurs.
Since interrupts are disabled on boot (on the MSP430, for instance), the user
needs to add code to their \texttt{INIT\_FUNC} to enable interrupts. However,
since interrupt handlers are assumed to be non-reentrant on our system, if
we are inside an interrupt handler, we need to keep interrupts disabled. We
also need to write the address of the task that we should return to upon boot -
the address of the \texttt{INTERRUPT\_TASK} on our system. If these are two
separate variables, there is a correctness issue with respect to power
failures. If we write the location of the \texttt{INTERRUPT\_TASK}, but die
before we can write the fact that we are in an interrupt handler, we will
return to te interrupt handler with interrupts enabled (since they will be
enabled by the \texttt{INIT\_TASK}. If we reverse the order and write the fact
that we are in an interrupt handler, then die before we can write the address
of the \texttt{INTERRUPT\_TASK}, we will resume execution at a non-interrupt
task with interrupts disabled, since the \texttt{INIT\_TASK} will not re-enable
them. Both of these cases could cause execution to fail; the former fails if
the interrupt handler is not written to be reentrant, the latter fails since
interrupts will never be processed. Thus, these writes need to be atomic.

@@

Instead, we exploit the fact that, generally, architectures are byte-aligned
with a fixed instruction size of 16 bits or more - this means, at the very
least, the lowest bit of any code address is zero. As this is true of ARM,
RISC-V, MSP430, and most RISC architecutres, we feel that this is a reasonable
assumption. We use this fact to store the flag that is set if execution is
inside an interrupt handler in the lowest bit of the address of the task that
we are currently executing (the task that execution will return to upon
reboot). Thus, we can use a single write to encode all the needed information,
giving atomicity. For simplicity for the programmer, upon an interrupt firing,
an \texttt{interrupt\_prologue()} is run that does this write. The address stored
in the current context is still the address of the user-defined
\texttt{INTERRUPT\_TASK}. In our model, this write is the guarentuee that the
system has recieved the interrupt - if power fails between when the interrupt
fires and this write occurs, the interrupt will be ignored. This could be fixed
by having a special capacitor in the power system to accomidate the write, but
assumptions about the power system, arguably, should not be a part of \chain.

To avoid the pathological stack corruption case described in the motivation for
interrupts, we keep a flag in non-volatile memory that is set if a reboot has
occured while in an interrupt handler and is unset otherwise. This flag is
unset in the afformentioned \texttt{interrupt\_prologue()} and set in
\texttt{main()}. If and only if the flag is unset when
\texttt{return\_from\_interrupt()} is called, the two values pushed on the
stack from an interrupt firing are popped off. Otherwise the stack remains
unchanged, since the values will have been effectively cleared by a power
failure.

@@

\section{Evaluation}
To test our multithreading library and mutex implementation, we instrumented
several existing Chain applications that do not share any tasks or channels to
run concurrently. The programs were run on either an msp430fr5969 launchpad or
on a Capybara sensing node, a custom intermittent development board with an
msp430fr5949 MCU. Reboot tests were carried out strictly on the Capybara board
because was simpler to guarantee the isolation of the MCU from the programming
fet than on the launchpad. We did not seekt to fully verify our implementation
of multi-threading, but we use our experience in application development to
support our claim that the threads behave correctly, and as a programmer would
expect.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.9\columnwidth]{figs/tapir}
\end{figure}

>>>>

In this paper we introduce the \textit{Threaded Application Programming on Intermittent Resources}
(TAPIR) model to expand the current intermittent programming model \textit{Chain} to
include general multiprogramming. TAPIR allows the range of possible programs for
intermittent devices to include more general computation functionality while leaving the
memory and energy overhead low enough that intermittent computation on those workloads
still guarantees forward progress. While nonvolatile memory usage will increase per thread
scheduled, certain optimizations are possible to reduce memory overhead overall. Additionally,
context switching through our scheduler task is a low-instruction count static overhead that is
easily accounted for in energy calculations. Through the design of TAPIR we show that general
purpose threaded applications are feasible on an intermittently powered device.
\end{abstract}

@@

Our multithreaded model is build on two existing programming models, DINO and
Chain.  An initial intermittent programming model DINO breaks intermittent
programs into tasks, which creates automatic checkpointing locations during
execution \cite{Dino}. DINO uses a checkpointing and recovery model that tracks
both volatile and nonvolatile memory.  The DINO execution model executes all
tasks transactionally, leading to atomic task semantics and does not permit
consistency violations in nonvolatile state \cite{Dino}. Chain, a more recent
iteration of the concepts in DINO, uses volatile state rarely, if ever, and uses
task boundaries called channels to store most used data in nonvolatile memory
\cite{Chain}. At compile time, Chain allocates the necessary channels between
tasks by parsing channel read and write statements in order to match up aligning
tasks. All inputs and outputs to each task, as well as necessary data
structures, are stored in a channel for a respective task. A channel may only be
modified by one task, and only one task may run at a time, which mitigates time
of check, time of use bugs (TOCTOU) in intermittence.  The one exception is a
self-channel for recursive code, which is double buffered.  Keeping all
modifiable data in nonvolatile memory that is not open to easy rewrite makes
these tasks idempotent across reboots.  Because of continuous reboots, many
operations necessarily will be executed more than once, and correctness is
compromised if those operations are not guaranteed to have the same outcome by
the programming model after each reboot.

@@

\section{Related Work}
In addition to the division of a program into tasks, as described above, prior work
explores several different strategies for preserving state across reboots. The approaches
range from fully independent of the programmer~\cite{ratchet, dewdrop} at one extreme, to programmer
defined task boundaries~\cite{Dino} at the other.  Compiler driven approaches such as the
method used in ~\cite{Mementos} allow the programmer to specify high level behavior of the
checkpointing system. A key similarity among the prior work is it
does not lend itself well to multi-programming let
alone multi-threading. In fact, many of these works state the assumption that the device
hardware cannot support a conventional OS, and thus focus on single threaded programs.
Despite the availability of (relatively) inexpensive nonvolatile memory technology such as
FRAM~\cite{quickrecall}, the energy and storage overhead of checkpointing multiple
thread states would limit the maximum per-thread checkpoint size.

@@

While prior work has stopped short of suggesting an operating system for intermittently
powered devices, a number of different works touch on the need for these devices to
perform multiple tasks in the same workload. Buettner, Greenstein and Wetherall motivate
their runtime, Dewdrop, based on difference in energy requirements for the tasks that a
CRFID needs to perform~\cite{dewdrop}. The energy aware memory mapping in ~\cite{Aware}
discusses the challenge of handling non-deterministic interrupts received by intermittent
sensing platforms when a peripheral must be serviced immediately and control flow deviates
from single threaded execution. From a hardware standpoint, the federated energy scheme in
~\cite{ufop} provides a platform for removing the energy dependence between an MCU and its
peripherals. Federated energy also simplifies a programmer's reasoning about reading from
a single peripheral. However, federating energy does not help a programmer describe how to
handle {\em multiple} peripherals. Consider a listing in
~\cite{ufop} reproduced in Figure ~\ref{fig:copy}, and imagine the level of understanding
of the hardware behavior a programmer would have to employ to, say, read one sensor if it
is available but read from and proceed to process data from a different sensor if it is not.

@@

The wireless sensor networks community has explored multiple approaches for handling
multitasking on resource limited devices. The designs tend to provide a combination of
timer and event driven context switching as well as a simple programming interface.
TinyOS uses a simple FIFO scheduler to handle event based context switches~\cite{tiny}.
Nano-RK implements priority based preemptive multitasking where task priorities are
statically defined~\cite{nano}. Contiki uses proto-theads, stackless program threads, to
offer a model between Nano-RK and tinyOS. The Contiki implements "cooperative
multithreading" such that events can trigger a context switch, but multithreading may also
be timer based~\cite{contiki}. The goal of multitasking interfaces in wireless sensor
networks is to provide a tradeoff between network lifetime and quality of service. This
differs from the intermittent programming model where the challenge is not system
lifetime, but system functionality based on energy availability.

@@

\section{System Implementation}
TAPIR has a round robin scheduling algorithm. At the end of a task, the compiler
inserts a jump back to the scheduler task, where the next task in the list is
the one to be executed. An alternate scheduling model that was considered was
a context switch on reboot instead of at the end of each task. The main drawback
of using a reboot-based scheduler is that, because power failure is the common
case, there is a high chance of reducing or halting forward progress altogether
because one thread holds a lock that all of the other threads need but cannot
complete. For that reason the design specifies end of task, so each task has
an opportunity to use a shared resource and release any locks before a power-
intensive context switch occurs. Additionally, a round robin scheduler was
chosen for this design because it is simple to implement and uses only one index counter,
very little power or memory overhead as compared to other scheduling algorithms. It is
critical for multithreading in Chain to not add significant power and memory
overhead in scheduling over conventional Chain so that the workload can
make as much forward progress as possible on one charge cycle. While out of
scope for this project, future work may include a priority scheduler to decrease
latency for certain threads due to context switching for I/O or other input latency
sensitive workloads.

When designing threading in Chain, the shared memory model is a main indicator
of the threaded execution model. The two most probable options when designing
TAPIR were a SIMD execution model of \textit{Thread Pools}, or a MIMD
traditional multiprogramming model. The Thread Pool design makes use of task
code duplication by explicitly forking a number of chain programs, all of which
run simultaneously on the scheduler on arrayed data. This option is useful
multiple data streams from I/O in embedded applications that need to be
processed before new data overrides the I/O device. Additionally, the Thread
Pool model makes sharing memory more regular by using multiple copies of the
same thread. However, this option is not as general purpose. The TAPIR design
ultimately settled on a multiprogramming model, because this allows for the
programmer to explicitly program unique chain programs with shared or discrete
data structures on the same system. However, this execution model requires a
more structured locking mechanism to avoid RMW during execution.

SIMD execution on an intermittent system relies on nonvolatile memory instead
of independent registers for each thread. In Chain's case the implementation
would rely on channels for each thread running in the system. In the MIMD
execution model channels could potentially be re-assigned to already allocated
channels because the threads may end at interleaving times. However, in the
SIMD model it is expected that each identical thread is at a comparable point
in execution to the other running threads, and so each thread that may be
run necessitates a new channel allocation. This could lead to an unavoidable
hard limit on the number of threads that can run in an application which in
turn limits potential applications that are useful on this intermittent device.

@@

\subsection{Locking Primitive}
To improve a programmer's control of shared data between threads, and to reduce
the complexity of accessing data that may be touched by many different tasks we
implement a simple mutex object that can be added to a channel. The mutex is
designed to be used in a standard \texttt{channel}, not a double buffered
\texttt{self\_channel}, so it must be able to tolerate a failure after a
read/write operation (i.e. test and set). To achieve idempotence, we add a
\texttt{thread\_id} to the \texttt{free} bit inside of each mutex. When
attempting to acquire a lock, a thread checks if its thread id is equal to the
lock holder, then it checks if the lock is free. If either condition is true,
the thread successfully acquires the lock, otherwise the \texttt{mutex\_lock}
function fails. Upon successfully acquiring the lock, a thread writes its id to
the mutex's holder id, and changes the \texttt{free} bit to \texttt{status =
held}. This scheme guarantees correctness because if a failure occurs after the
mutex is set but before the task commits, the task will read out the id of the
mutex holder before checking if it is free. Upon matching its id with the holder
id, the task will proceed regardless of the fact that \texttt{status = held}.
%TODO Maybe add a figure here to make the sequence of read/write/failure clearer...
Further, a task must finish executing before switching to another thread, so
there is no possibility for an external thread to see a value that was not
committed by another task.

It is important to note that the mutex object can be modified idempotently, but
accesses to the data in the same channel as the mutex must adhere to channel
exclusion. In our example application, the mutex is contained within a user
defined \texttt{shared\_channel} structure that is analogous to a
\texttt{multicast\_channel} but abandons the naming convention that "restricts"
the channel to a static subset of tasks. Any task may read {\em or} or write to
the \texttt{shared\_channel}, but performing both would break correctness. In
the future, a compiler could be used to guarantee that a
\texttt{shared\_channel} is not read and written by the same task, but at
present we believe that the \texttt{mutex} and \texttt{shared\_channel}
interface is easier for programmers to use than \texttt{multicast\_channels}, or
a specialized double buffered channel when passing data between threads. It
maintains familiar semantics for programmers and does not incur the memory
overhead of double buffering.

@@

\begin{figure}[ht]
\begin{minipage}[b]{.9\columnwidth}
  \includegraphics[width=0.7\columnwidth]{figs/int_ctrl_flow}
  \caption{Interrupt Control Flow}\label{intCtrlFlow}
\end{minipage}
\end{figure}

@@

\section{Implementation}
\subsection{Scheduler}

The data structures for our scheduler are listed in figure \ref{schDataStruct}.
Like the rest of chain, memory for data structures is allocated at compile
time. Thus, the limit on the maximum number of threads allowed in a system is
fixed arbitrarily - not ideal, but there isn't a better option while operating
inside chain. The \texttt{threads} array is stored in all-to-one channel, from
every task to the scheduler task. The \texttt{free\_indices} array is stored
in a one-to-all channel, from the scheduler task to every task.

With these multicast channels, the scheduler task can act as a buffer between
multiple threads - arbitrary tasks write \texttt{threads[]} with calls to
\texttt{thread\_create} and \texttt{thread\_end} and tasks read from
\texttt{free\_indices} through calls to thread create, maintaining
idempotency. Inside the scheduler task, we read and write \texttt{current},
so \texttt{current} is double buffered. The only issue is with
\texttt{thread\_init} since it is called from user-defined tasks and
initializes \texttt{current} - the same task that called \texttt{thread\_init}
could call \texttt{thread\_create}, and the \texttt{thread\_create} call
wouldn't see the update to the variable. Since \texttt{current} is always
initialized to the same value (0), we opt for manually swapping the buffer on
\texttt{current} so that other threads can see the update.

\begin{figure}[ht]
\begin{minipage}[b]{.9\columnwidth}
\begin{lstlisting}[language=C]
    typedef struct thread_t {
        unsigned thread_id;
        context_t context;
    } thread_t;

@@

    // List of free indices in threads[]
    unsigned free_indices[MAX_NUM_THREADS];
    unsigned free_indices_size;

    // Index of current running thread
    // Stored in self channel
    unsigned current;
\end{lstlisting}
\caption{Scheduler Data Structures}\label{schDataStruct}
\end{minipage}
\end{figure}

We keep an unsigned integer, \texttt{curr\_free\_index}, in volatile memory,
zeroed on every reboot, that acts as a current pointer in
\texttt{free\_indices[]}. Every time the scheduler task runs,
\texttt{free\_indices[]} is updated with the state of \texttt{threads[]},
based on whether or not the active bit is set for a given index. When
\texttt{thread\_create} is called, if \texttt{curr\_free\_index} is greater
than \texttt{free\_indices\_size}, then \texttt{thread\_create} will fail in
POSIX fashion - the system does not have enough resources to support another
thread. Otherwise, \texttt{threads[free\_indices[curr\_free\_index]]} is set
to the new thread. Upon power failure, \texttt{curr\_free\_index} is reset to
zero, so as long as the rest of the operations in the task are idempotent, so
will \texttt{thread\_create} calls - the same locations in \texttt{threads[]}
will be overwritten.

\begin{figure}[ht]
\begin{minipage}[b]{.9\columnwidth}
  \includegraphics[width=0.7\columnwidth]{figs/threadsArray}
  \caption{Threads Array}\label{threadsArray}
\end{minipage}
\end{figure}

\subsection{Interrupts}
The primary issue with interrupts is the interrupt enable flags. The MSP430,
like many other embedded architectures has a master interrupt enable,
as well as specific, device-level interrupt enables. Both are the user's
responsibility, and need to be specified in the user-defined
\texttt{INIT\_FUNC}, as both will be cleared on reset. When an interrupt fires,
however, interrupts will be disabled by the CPU (again, on most architectures)
and if the device loses power immediately after an interrupt occurs, there will
be no record that the interrupt occurred. This case is unavoidable on
intermittent power without explicit support from the power system - we reason
that if the user cared about every interrupt that fired and if missing one
interrupt would be catastrophic to the device's operation, then, arguably, that
device should not be intermittently powered.

Beyond this, there remains the issue of how to return to the interrupt task
with the interrupt enable set to the correct value if a power failure occurs.
Since interrupts are disabled on boot (on the MSP430, for instance), the user
needs to add code to their \texttt{INIT\_FUNC} to enable interrupts. However,
since interrupt handlers are assumed to be non-reentrant on our system, if
we are inside an interrupt handler, we need to keep interrupts disabled. We
also need to write the address of the task that we should return to upon boot -
the address of the \texttt{INTERRUPT\_TASK} on our system. If these are two
separate variables, there is a correctness issue with respect to power
failures. If we write the location of the \texttt{INTERRUPT\_TASK}, but die
before we can write the fact that we are in an interrupt handler, we will
return to the interrupt handler with interrupts enabled (since they will be
enabled by the \texttt{INIT\_TASK}. If we reverse the order and write the fact
that we are in an interrupt handler, then die before we can write the address
of the \texttt{INTERRUPT\_TASK}, we will resume execution at a non-interrupt
task with interrupts disabled, since the \texttt{INIT\_TASK} will not re-enable
them. Both of these cases could cause execution to fail; the former fails if
the interrupt handler is not written to be reentrant, the latter fails since
interrupts will never be processed. Thus, these writes need to be atomic.

@@

Instead, we exploit the fact that, generally, architectures are byte-aligned
with a fixed instruction size of 16 bits or more - this means, at the very
least, the lowest bit of any code address is zero. As this is true of ARM,
RISC-V, MSP430, and most RISC architectures, we feel that this is a reasonable
assumption. We use this fact to store the flag that is set if execution is
inside an interrupt handler in the lowest bit of the address of the task that
we are currently executing (the task that execution will return to upon
reboot). Thus, we can use a single write to encode all the needed information,
giving atomicity. For simplicity for the programmer, upon an interrupt firing,
an \texttt{interrupt\_prologue()} is run that does this write. The address stored
in the current context is still the address of the user-defined
\texttt{INTERRUPT\_TASK}. In our model, this write is the guarantee that the
system has received the interrupt - if power fails between when the interrupt
fires and this write occurs, the interrupt will be ignored. This could be fixed
by having a special capacitor in the power system to accommodate the write, but
assumptions about the power system, arguably, should not be a part of \chain.

To avoid the pathological stack corruption case described in the motivation for
interrupts, we keep a flag in non-volatile memory that is set if a reboot has
occurred while in an interrupt handler and is unset otherwise. This flag is
unset in the aforementioned \texttt{interrupt\_prologue()} and set in
\texttt{main()}. If and only if the flag is unset when
\texttt{return\_from\_interrupt()} is called, the two values pushed on the
stack from an interrupt firing are popped off. Otherwise the stack remains
unchanged, since the values will have been effectively cleared by a power
failure.

@@

\section{Evaluation}
To test our multithreading library and mutex implementation, we instrumented
several existing Chain applications that do not share any tasks or channels to
run concurrently. The programs were run on either an MSP430FR5949 launchpad or
on a Capybara sensing node, a custom intermittent development board with an
MSP430FR5949 MCU. Reboot tests were carried out strictly on the Capybara board
because was simpler to guarantee the isolation of the MCU from the programming
FET than on the launchpad. We did not seek to fully verify our implementation
of multi-threading, but we use our experience in application development to
support our claim that the threads behave correctly, and as a programmer would
expect.

@@

\begin{figure}[hb]
  \centering
  \includegraphics[width=0.9\columnwidth]{figs/tapir}
\end{figure}
